\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-11-02 demystify.tex
add stuff for nonCS/nonSTEM

\begin{document}
\setcounter{section}{-1};

\maketitle

@@

Yes, as long as you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \S\ref{sec:improve-your-chance}.

@@

\subsection{Difference between PhD in the US and other countries}
Here I try to list the main differences between CS PhD in the US and other countries.  This is based on my experience in the US and what I heard from others who did PhD in other countries.

@@

\emph{Project proposal}: in many countries, you have to choose a project and advisor \emph{during} the application process (e.g., you write a proposal for potential advisor). In the US, you start your PhD without a project and advisor and find them later. This is also why US PhD programs are longer.

@@

\emph{Funding}:  In Europe, stipend (salary) comes from the university or from Gov research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your advisor (and no fixed duration).  There are fewer TA opportunities in the European universities compared to the US.

\emph{Faculty Position after PhD}: In other countries, PhD graduates typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates interested in academia can apply directly for faculty position.  This is why US PhD programs are longer and more focused on research.

@@

      \begin{center}
        \includegraphics[scale=0.5]{c6.png}
      \end{center}

@@

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for funding).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student makes two conference trips per year, one domestic and one international (so conf. registration, airline tickets, taxi, meals, etc are all included). At the end, the total budget comes out to be \$69K. \emph{The summary is that while you're paid X, your advisor probably pays 2X for you}.

@@

\subsection{Will I be miserable during my a PhD?}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable.  Sure there are issues with PhD and academic in general, but then there are also issues any jobs, family, and life.  Certainly there are horrible professors, bad relationships with co-authors and officemates, feeling left out or discriminated (e.g., because you're an international student), and it is good to be aware of those.  However, in reality there are many good mentors, supportive labmates and department, happy students, etc.  Don't let social media make you feel pessimistic and destroy your quest to advance knowledge.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that anyone can contribute to it.

>>>>

\begin{document}

\maketitle
\setcounter{section}{-1};

@@

Yes, \emph{as long as} you can demonstrate you are ready for CS PhD research through research experiences, LoRs, statements, etc as mentioned. You might be even able to leverage this to make your profile stand out as mentioned in \S\ref{sec:improve-your-chance}.

One main concern for non-CS or non-STEM students is if you have the sufficient technical background, typically obtained through core CS courses.  So you will want to show that you have such knowledge through your coursework, projects, or research. 
For example, if you have taken a course on Algorithms, even online ones like Coursera, you can talk about it in your statement.  If you have done a project or have a professional certification (e.g., A+) through work that requires knowledge of OS, you can talk about it in your statement.  If you have done research that requires knowledge of Discrete Maths, you can talk about it in your statement.  You can also ask your LoR writers to talk about your technical background.  
In summary, in your application, convince us that you have the technical background to do CS PhD research. 
Note that many online courses teaching about AI does not really help much as these are not core CS knowledge.

In short, you \emph{do not need} to formally taking CS courses, you just need to show that you have these essential knowledge through ways as mentioned. Many universities are well-aware that incoming graduate students might not have all the technical background, so they often have a "bridge" courses to help students catch up.  For example, GMU has 4 bridge courses (Data Structure and Algorithm, Computer Systems, Discrete Maths, and Programming Foundation) that incoming students can take to catch up on their CS knowledge.

\begin{commentbox}[Vu:]
    I would strongly advocate to admit a student with a non-STEM background (e.g., business, English major) but can show that they have a strong drive for CS by studying core CS knowledge through various channels (e.g., self-study through online courses, projects, etc).  I have seen many students with non-CS background who are very successful in CS PhD.  In fact, I have seen many students with CS background who are not successful in CS PhD.  So it is not about your background, it is about your drive and passion for CS research.    \end{commentbox}

@@

\subsection{Difference between PhD in the US and other countries}
Here we summarize the main differences between CS PhD in the US and other countries.  This is based on my experience and what I heard from others who did PhD in other countries.

@@

\emph{Project proposal}: in many countries, you have to choose a project and advisor \emph{during} the application process (e.g., you write a proposal to a potential advisor). In the US, you often start your PhD without a project and advisor and find them later. This is also why US PhD programs are longer.

@@

\emph{Funding}:  In Europe, stipend (salary) comes from the university or from gov't research organizations and often have a fixed duration (e.g., 3 years).  In the US, stipend (e.g., RA) comes directly from your advisor (no fixed duration).  There are also fewer TA opportunities in the European universities compared to the US.

\emph{Faculty position after PhD}: In other countries, PhD graduates typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates interested in academia can apply directly for faculty position.  This is why US PhD programs are longer and more focused on research.

@@

      \begin{center}
        \includegraphics[scale=0.3]{c6.png}
      \end{center}

@@

Tab.~\ref{tab:cost} shows the budget breakdown for a GRA per year (this level of details is what faculty actually uses when applying for grants).
These numbers are based on my experience at public universities in the US.  Private universities may have different numbers.  For simplicity, I will assume the department has a 9-month stipend of \$27000 (GMU actually pays more) and therefore a 3-month summer of \$9000. I will also use GMU tuition rate of about \$15,000/year for full-time study (which is quite cheap compared to private universities, e.g., MIT charges around \$50K) and a 58.9\% rate on \emph{indirect cost}, which is what GMU charges for overhead or administrative costs (yes, after all, universities are businesses!).  Finally, I assume the student take two conference trips per year, one domestic and one international (conf. registration, airline tickets, taxi, meals, etc are all included).

At the end, the total budget comes out to be \$69K/year to support a PhD student. \emph{The summary is that over your 5-6 year of your PhD, you cost about \$350K--400K, and while you're paid X, your advisor probably pays 2X for you}.

@@

\subsection{Will I be miserable during my a PhD?}\label{sec:miserable}
Many posts from VietPhD, a Facebook group for Vietnamese PhD students, are about how PhD students are mistreated, stressed, and miserable.  Indeed there are issues with PhD and academic in general, but then there are also issues with jobs, family, and life.  Certainly there are bad relationships with professors, co-authors, and officemates, and toxic working environment, feeling left out or discriminated (e.g., because you're an international student), and it is good to be aware of those.  However, in reality there are many good mentors, supportive labmates and department, happy students, etc.  So don't let social media make you feel pessimistic and destroy your quest to advance knowledge.

@@

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). I have put the source code of this document on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that anyone can contribute to it.
